Computing Direct Sum Decompositions

### 2025-04-22 paper/main.tex
changes before fixing a mistake

\let\oldb\b
\let\oldl\l
\usepackage[dvipsnames]{xcolor}
\usepackage{notes,hyperref}
\usepackage{mathtools}
\usepackage{cleveref}
\let\b\beta
\let\l\lambda
\let\wtilde\widetilde
\let\bar\overline
\def\sing{{\mathrm{sing}}}

@@

\begin{abstract}
  %The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, representation theory, topological data analysis, and other fields.
  \devlin{We had the commented-out sentence in both the abstract and intro; I think we can just remove it from the abstract? If not, at least rephrase}
  The purpose of this work is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$: the first algorithm applies in the (multi)graded case, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space); the second algorithm applies when $R$ is local and $k$ is a finite field, which opens the door to computing decompositions in singularity theory. We also present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards (including in the non-graded case) and syzygies over Artinian rings.
\end{abstract}

\section{Introduction}
\label{sec:intro}

The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, group theory, representation theory, and other fields. Within commutative algebra, for instance, the classification of rings $R$ for which there are only finitely many isomorphism classes of indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems (see, for example, \cite{Yoshino90,LW12} for the former and \cite{SVdB97,Hara15,TT08}, among many others, for the latter). For both these problems, and many others, making and testing conjectures depends on computing summands of modules and verifying their indecomposability.
The need for, and absence of, an algorithm to do so is noted, for example, in
\cite[15.10.9]{Eisenbud95}.

Currently there are no efficient \devlin{would it be too bold to remove ``efficient''?} algorithms available for checking indecomposability or finding summands of arbitrary modules over commutative rings. In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite-dimensional modules over a group algebra have wide ranging applications in computational group theory \cite{Parker84,HR94,Holt98} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.

@@

The purpose of this paper is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$, when either $R$ is (multi)graded and $k$ an arbitrary field, or $R$ local and $k$ a field of positive characteristic.
In the graded case, our algorithm is often capable of completely decomposing a module in a single pass, while in the local case it proceeds by recursively finding summands and then reapplying the algorithm to each summand.
%We also describe applications of our algorithm for computations in characteristic 0.
The multigraded case of our algorithm enables the computation of indecomposable summands of coherent \emph{sheaves} on subvarieties of toric varieties (in particular, for varieties embedded in $\PP^n$).

After describing the algorithms and proving their correctness, we present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings. In particular, we highlight the results of \cite{CDE24}, which shows a recurrence formula for indecomposable summands of high syzygies of the residue field of Golod rings, made possible through experiments and observations using our algorithm.

@@

\begin{rem}
  There is a ``missing'' case of our algorithm: when $R$ is only local, not graded, and $k$ has characteristic 0.
  If a module over a ring of characteristic 0 is decomposable, its reductions modulo $p$ will be as well; thus, our algorithm at least provides a heuristic for verifying decomposability in characteristic 0. See \Cref{rem:char}.
  %Although the local algorithm described below is only proved to result (probabilistically) in a decomposition into indecomposable summands in positive characteristic, in practice it often does produce nontrivial indecomposable decompositions even in characteristic 0.
  %Although the algorithm described below is only proved to result (probabilistically) in a decomposition into indecomposable summands in positive characteristic, in practice it often does produce nontrivial indecomposable decompositions even in characteristic 0. Moreover, if a module over a ring of characteristic 0 is decomposable, its reductions modulo $p$ will be as well; thus, our algorithm provides a heuristic for verifying decomposability in characteristic 0.

@@

\subsection*{Acknowledgements}
The authors would like to thank David Eisenbud for several very useful conversations and ideas, and Ezra Miller for pointing out the relevance of this algorithm in topological data analysis.
We also thank the American Institute of Mathematics for hosting the ``Macaulay2: expanded functionality and improved efficiency'' workshop, during which some of this work took place.
The work of the first author was supported in by the National Science Foundation RTG Grant No.~1840190.
The second author was supported in part by the Doctoral Dissertation Fellowship at the University of Minnesota and the National Science Foundation under Grant No.~2001101.
We also thank the referees for their careful reading and suggestions, which significantly improved the exposition of the paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Notation}
\label{sec:notation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The main algorithms}
\label{sec:main-algs}

@@

\subsection{The graded algorithm}\label{sec:main-alg-new}

Let $R$ be a $\N$-graded ring, with $R_0 = k$, an algebraically closed field.
%Assume furthermore that the degree support of $R$ is strongly convex, i.e., that
%%the effective cone
%$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
Let $M$ be a finitely generated, graded $R$-module and consider a degree zero endomorphism $\phi \in [\End M]_0$.

@@

\begin{lem}\label{graded_algorithm}
  Let $N = \mu(M)$ and set $\psi = \phi^N$. Then $M = \im\psi \oplus \coker\psi$.
\end{lem}
\begin{proof}
  Fix homogeneous generators $m_1,\dots, m_n$ of $M$, with $\deg m_i$ nonincreasing.
  The matrix for $\phi$ is then ``block upper triangular'' with diagonal blocks with entries in $k$, i.e., of the form
  \[ \phi = \begin{pmatrix}
    B_1    & *      & \dots  & *      \\
    0      & B_2    & \dots  & *      \\
    \vdots & \vdots & \ddots & \vdots \\
    0      & 0      & \dots  & B_l
  \end{pmatrix}, \]
  where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $R$. To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$.

  Thus, since each matrix $B_i$ has entries in the algebraically closed field $k$, by performing row and column operations over $k$ we can put each $B_i$ into Jordan canonical form $J_i$, with each Jordan block having entries in $k$. This corresponds to replacing each generator $m_i$ with a $k$-linear combination of generators $m_j$ with $\deg m_i = \deg m_j$, which is an endomorphism of $M$. By doing so starting with $B_1$ and proceeding to $B_l$, the resulting matrix has the form
  \begin{equation}
    \label{eq:Jordan}
    \phi
    =
    \begin{pmatrix}
      J_1    & *      & \dots  & *      \\
      0      & J_2    & \dots  & *      \\
      \vdots & \vdots & \ddots & \vdots \\
      0      & 0      & \dots  & J_l
    \end{pmatrix},
  \end{equation}
  where each $J_i$ is a Jordan block over $k$ with eigenvalue $\l_i$, and the remaining entries are arbitrary.
  Relabeling, we may assume that $\l_i\neq 0$ for $i=1,\dots,l_0$ and $\l_i=0$ for $i=l_0+1,\dots,l$.

  If we take $N_0  $ to be at least the sum of the sizes of the nilpotent Jordan blocks $J_{l_0+1},\dots,J_l$, then for $N\geq N_0$ we have
  $$
  \phi^N =
  \begin{pmatrix}
    J_1^N & *** &\dots & \dots& \dots  &\dots & *** \\
    0 & J_2^N & \dots &\dots&\dots & \dots& *** \\
    \vdots &\vdots & \ddots &\vdots \\
    0 & 0 & \dots & J_{l_0}^N & *** & \dots & ***\\
    0 & 0 & \dots & 0 & 0 & \dots & 0\\
    \vdots &\vdots & \ddots &\vdots \\
    0 & 0 & \dots & 0 & 0 & \dots & 0
  \end{pmatrix},
  $$
  where each nonzero $J_i^N$ has the units $\l_i^N$ on the diagonal.
  ($\mu(M)$ is an upper bound on the sum of the sizes of the nilpotent Jordan blocks, so the above holds for $N\geq \mu(M)$.)

@@

  So, all that needs to be seen is that $\psi:=\psi^N$ is surjective.
  To see this, note that after replacing the original generators for $M$ by the generators that put $\psi$ into the form above,  a generating set for $\im(\phi^N)$ is given by $f_i = \psi(m_i) = u_i\cdot e_i + \sum_{j<i } a_{ij} e_j$ for $i = 1,\dots,n-N_0$, where $u_i$ is a unit (explicitly, $u_i$ is a power of some nonzero $\l_{i'}\in k$).
  That is, a generating set is given by
  $$u_1 e_1,\quad  u_2 e_2+a_{21}e_1,\quad \dots ,\quad u_{n-n_0}e_{n-N_0}+a_{n-N_0,n-1}e_{n-N_0-1}+\dots + a_{N-n_0,1}e_1.$$
  Since the $u_i$ are units, the $R$-span of these generators is clearly the same as the $R$-span of $e_1,\dots,e_{n-N_0}$, and thus after applying $\psi$ we get back the $R$-span of $f_1,\dots,f_{n-N_0} $, which is $\im \psi^N$. Thus, $\phi: \im \psi^N \to \im \psi^N$ is surjective, and the result follows.
\end{proof}

@@

The above can of course produce the trivial direct sum decomposition: if the eigenvalues of $\phi$ are all nonzero, the proof yields that $\im \psi = M$ and $\coker \psi = 0$. The following corollary of the proof of \cref{graded_algorithm} ensures that we can always assume $\phi$ has zero as an eigenvalue:

\begin{lem}
  The eigenvalues of $\phi$ lie in $k$.
  In particular,
  for any  $\phi \in \End M$, there is $\l \in k$ such that $\phi - \l \id_M \in \End M$ has zero as an eigenvalue.
\end{lem}

@@

\begin{algm}
  \label{alg:graded}
  Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
  \begin{enumerate}
  \item\label{item:End0} Take a random element $\phi_0$ of $[\End_R(M)]_0$, the degree zero part of $\End_R(M)$.
  \item For an eigenvalue $\mu $ of $\phi_0$, let $\phi=\phi_0-\mu \id_M$.
  \item If $\phi$ has at least two eigenvalues, let $\psi=\phi^{\mu(M)}$. By \Cref{graded_algorithm}, $M\cong \im \psi \oplus \coker \psi$.  \item Repeat steps (1)--(3) for both $\im \psi$ and $\coker \psi$.
  \end{enumerate}
\end{algm}

@@

\subsection{The local algorithm}\label{sec:main-alg}

@@

\begin{lem}\label{lem:jordan}
  Let $k$ be a finite field of characteristic $p$, and let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$. If $\l$ is an eigenvalue of $A$, then some power of $A-\l I$ is idempotent.
  Furthermore, if $\l$ is not the only eigenvalue of $A$, then a  nonzero power of $A-\l I$ is idempotent.
\end{lem}
\begin{proof}
  Since all eigenvalues of $A$ are contained in $k$, we can without loss of generality put $A$ in Jordan canonical form, with each Jordan block having the form
  \[ r_i\Biggl\{
  \underbrace{\begin{pmatrix}
      \l_i & 1 & 0 & \dots & 0 \\
      0 & \l_i & 1  & \dots & 0 \\
      \vdots & & & & \vdots \\
      0 & 0 & 0 & \dots & \l_i
  \end{pmatrix}}_{r_i} \]
  with each $\l_i$ an eigenvalue of $A$.
  In this basis, $A-\l I$ will be block-diagonal with blocks of form
  \[ \begin{pmatrix}
    \l_i-\l & 1 & 0 & \dots & 0 \\
    0 & \l_i-\l & 1 & \dots & 0 \\
    \vdots & & & & \vdots \\
    0 & 0 & 0 & \dots & \l_i-\l
  \end{pmatrix}. \]
  Set $\mu_i=\l_i-\l$. Then for any $l\geq 1$, the $l$-th power $(A-\l I)^l$ is block-diagonal with blocks of the form
  \[ \begin{pmatrix}
    \mu_i^l & \binom{l}1\mu_i^{l-1} & \binom{l}2\mu_i^{l-2} & \dots & \binom{l}{r_i}\mu_i^{l-r_i} \\
    0 & \mu_i^l & \binom{l}1\mu_i^{l-1} & \dots & \binom{l}{r_i-1}\mu_i^{l-r_i+1} \\
    \vdots & & & & \vdots \\
    0 & 0 & 0 & \dots & \mu_i^l
  \end{pmatrix}. \]
  If we choose $l > r_i$ and $l$ a power of $p$, say $p^{e_0}$, then all non-diagonal terms will vanish, so all blocks will have the form
  \[ \begin{pmatrix}
    \mu_i^l & 0 & 0 & \dots & 0 \\
    0 & \mu_i^l & 0 & \dots & 0 \\
    \vdots & & & & \vdots \\
    0 & 0 & 0 & \dots & \mu_i^l
  \end{pmatrix}. \]
  Finally, since $k\subset \FF_{p^e}$ for some $e$, if we choose $l$ to be divisible also by $p^e-1$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1.$$
  In particular, taking $l=p^{e_0}(p^e-1)$,
  $(A-\l I)^{p^{e_0}(p^e-1)}$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent. Note moreover that if some $\l_i\neq \l$, then $(A-\l I)^l$ is not the zero matrix.
\end{proof}

@@

\begin{algm}\label{alg:local}
  Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
  \begin{enumerate}
    %In the graded case, take a general element $A_0$ of $[\End_R(M)]_0$, the degree zero part of $\End_R(M)$; in the local case,
  \item Take a general element $A_0$ of $\End_R(M) \setminus \m \End_R(M)$.
  \item Consider the resulting endomorphism $A$ of the $k$-vector space $M/\m M$ and find the eigenvalues of $A$.
  \item If $A$ has at least two eigenvalues, choose one eigenvalue $\l$, and compute a sufficiently high power $B=A^{p^{e_0}(p^{e}-1)}$ of $A$ (with the power explicitly as in the proof of \Cref{lem:jordan}). This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im B \oplus \coker B$ by \Cref{lem:idemp}.
  \item Repeat steps (1)--(3) for both $\im B$ and $\coker B$.
  \end{enumerate}
\end{algm}

\subsection{Termination of the algorithms}

@@

\begin{lem}
  \label{lem:term}
  Assume that either $(R,\m)$ is a graded ring with $R_0=k$, or $(R,\m)$ is a local ring with $k=R/\m$ a field of positive characteristic, and that $M$ is a finitely generated $R$-module (homogeneous in the graded case).
  If $M$ has an endomorphism with at least two distinct eigenvalues modulo $\m$, then $M$ is not indecomposable.
\end{lem}

@@

\begin{lem}
  \label{lem:distinct}
  Retain the notation of \Cref{lem:term}.
  \begin{itemize}
  \item If $R$ is graded and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero has at least two distinct eigenvalues (necessarily in $k$)
  \item If $R$ is local and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
  \end{itemize}
\end{lem}

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section{Decomposing coherent sheaves}
\label{sec:coherent}

@@

\begin{dfn}
  For $f\in \End_{\OO_X}(E)$, taking top exterior powers yields the map
  $$ \bigwedge^{\rank E} f: \det E  \to \det E, $$
  where $\det E := \bigwedge^{\rank E} E$.
  Since $\End_{\OO_X}(\det E) = H^0(\OO_X)=k$, $\bigwedge^{\rank E} f$ is multiplication by $\l\in k$; we write $\det f$ for this scalar $\l$.
\end{dfn}

@@

\begin{dfn}
  $\l \in k$ is an \defn{eigenvalue} of $f \in \End_{\OO_X}(E)$ if $\det(f-\l \id_E)= 0$. In other words, the eigenvalues of $f$ are the zeroes of the univariate polynomial $\det(f-t\id _E)$ with coefficients in $k$.
\end{dfn}

>>>>

\let\oldb\b
\usepackage[dvipsnames]{xcolor}
\usepackage{notes,hyperref}
\usepackage{mathtools}
\usepackage{cleveref}
\let\b\beta

\let\wtilde\widetilde
\let\bar\overline
\def\sing{{\mathrm{sing}}}

@@

\begin{abstract}
  We describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$: the first algorithm applies in the (multi)graded case, which enables the computation of indecomposable summands of coherent sheaves on subvarieties of toric varieties (in particular, for varieties embedded in projective space); the second algorithm applies when $R$ is local and $k$ is a finite field, which opens the door to computing decompositions in singularity theory. We also present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards (including in the non-graded case) and syzygies over Artinian rings.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

The problems of finding isomorphism classes of indecomposable modules with certain properties, or determining the indecomposable summands of a module, are ubiquitous in commutative algebra, representation theory, topological data analysis, and other fields. Within commutative algebra, for instance, the classification of rings $R$ for which there are only finitely many isomorphism classes of indecomposable maximal Cohen--Macaulay $R$-modules (the \emph{finite CM-type} property), or determining whether iterated Frobenius pushforwards of a Noetherian ring in positive characteristic have finitely many isomorphism classes of indecomposable summands (the \emph{finite F-representation type} property) are two well-established research problems (see, for example, \cite{Yoshino90,LW12} for the former and \cite{SVdB97,Hara15,TT08}, among many others, for the latter). For both these problems, and many others, making and testing conjectures depends on computing summands of modules and verifying their indecomposability.
The need for, and absence of, an algorithm to do so is noted, for example, in
\cite[15.10.9]{Eisenbud95}.

Currently there are no algorithms available for checking indecomposability or finding summands of arbitrary modules over commutative rings. In contrast, variants of the ``Meat-Axe'' algorithm for determining irreducibility of finite-dimensional modules over a group algebra have wide ranging applications in computational group theory \cite{Parker84,HR94,Holt98} and are available through symbolic algebra software such as Magma and GAP \cite{MAGMA,GAP}.

@@

The purpose of this paper is to describe and prove correctness of two practical algorithms for computing indecomposable summands of finitely generated modules over a finitely generated $k$-algebra $R$, when either $R$ is (multi)graded and $k$ an arbitrary field, or $R$ local and $k$ a field of positive characteristic.
In the graded case, our algorithm is often capable of completely decomposing a module in a single pass, while in the local case it proceeds by splitting the module into two summands and continuing recursively.
%We also describe applications of our algorithm for computations in characteristic 0.
The multigraded case of our algorithm enables the computation of indecomposable summands of coherent \emph{sheaves} on subvarieties of toric varieties (in particular, for varieties embedded in $\PP^n$).

Finally, we present multiple examples, including some which present previously unknown phenomena regarding the behavior of summands of Frobenius pushforwards and syzygies over Artinian rings. In particular, we highlight the results of \cite{CDE24}, which shows a recurrence formula for indecomposable summands of high syzygies of the residue field of Golod rings, made possible through experiments and observations using our algorithm.

@@

\begin{rem}
  There is a ``missing'' case of our algorithm: when $R$ is only local, not graded, and $k$ has characteristic 0. If a module over such a ring is decomposable, its reductions modulo $p$ will be as well; thus, our algorithm at least provides a heuristic for verifying decomposability in characteristic 0. See \Cref{rem:char}.
  %Although the local algorithm described below is only proved to result (probabilistically) in a decomposition into indecomposable summands in positive characteristic, in practice it often does produce nontrivial indecomposable decompositions even in characteristic 0.
  %Although the algorithm described below is only proved to result (probabilistically) in a decomposition into indecomposable summands in positive characteristic, in practice it often does produce nontrivial indecomposable decompositions even in characteristic 0. Moreover, if a module over a ring of characteristic 0 is decomposable, its reductions modulo $p$ will be as well; thus, our algorithm provides a heuristic for verifying decomposability in characteristic 0.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Acknowledgements}

The authors would like to thank David Eisenbud for several very useful conversations and ideas, and Ezra Miller for pointing out the relevance of this algorithm in topological data analysis.
We also thank the American Institute of Mathematics for hosting the ``Macaulay2: expanded functionality and improved efficiency'' workshop, during which some of this work took place.
The work of the first author was supported in by the National Science Foundation RTG Grant No.~1840190.
The second author was supported in part by the Doctoral Dissertation Fellowship at the University of Minnesota and the National Science Foundation under Grant No.~2001101.
We also thank the referees for their careful reading and suggestions, which significantly improved the exposition of the paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Notation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The main algorithms}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The graded case}\label{sec:graded-alg}

Let $R$ be a $\N$-graded ring, with $R_0 = k$, an algebraically closed field.
%Assume furthermore that the degree support of $R$ is strongly convex, i.e., that
%%the effective cone
%$\Eff R = \{ \deg r : r\in R\}$ is a strongly convex cone in $\ZZ^n$.
Suppose $M = \bigoplus M_d$ is a finitely generated, graded $R$-module and consider a degree zero endomorphism $\phi \in [\End M]_0$.

@@

\begin{lem}\label{lem:graded_algorithm}
  Let $N = \mu(M)$ and set $\psi = \phi^N$. Then $M = \im\psi \oplus \coker\psi$.
\end{lem}
\begin{proof}
  Fix homogeneous generators $m_1,\dots,m_N$ of $M$, with $\deg m_i$ nonincreasing, and let $d_1,\dots,d_n$ be the decreasing list of unique degrees of generators of $M$. The matrix for $\phi$ is then ``block upper triangular'' with $n$ diagonal blocks with entries in $k$, i.e., of the form
  \[ \phi = \begin{pmatrix}
    B_1    & *      & \dots  & *      \\
    0      & B_2    & \dots  & *      \\
    \vdots & \vdots & \ddots & \vdots \\
    0      & 0      & \dots  & B_n
  \end{pmatrix}, \]
  where the blocks $B_i$ are square matrices \emph{with entries in $k$}, and the remaining entries are in $R$. To see this, note that since $\phi$ has degree zero, $\phi(m_i)$ is a $k$-linear combination of the $m_j$ with $\deg m_j = \deg m_i$ plus an $R$-linear combination of the $m_j$ with $\deg m_j < \deg m_i$. In particular, $B_i$ is the induced endomorphism on the submodule generated by all generators of degree $d_i$.

  Since the matrices $B_i$ have entries in the algebraically closed field $k$, by performing row and column operations over $k$ we can put each $B_i$ into Jordan canonical form $J_i$ with entries in $k$. In particular, since the generators of degree $d_i$ are linearly independent, a $k$-linear change of basis on this set induces an automorphism $\alpha_i$ of $M$.

  Let $\alpha$ be the composition of these automorphisms for each block $B_i$, then $\phi' = \alpha^{-1}\circ\phi\circ\alpha$ corresponds to performing row and column operations on $\phi$ over $k$ and has the form
  \begin{equation}
    \label{eq:Jordan}
    \phi' = \begin{pmatrix}
      J_1    & *      & \dots  & *      \\
      0      & J_2    & \dots  & *      \\
      \vdots & \vdots & \ddots & \vdots \\
      0      & 0      & \dots  & J_n
    \end{pmatrix},
  \end{equation}
  where each $J_i$ is a Jordan block over $k$ with eigenvalue $\lambda_i$, and the remaining entries are in $R$.
  Relabeling, we may assume that $\lambda_i\neq 0$ for $i=1,\dots,n_0$ and $\lambda_i=0$ for $i=n_0+1,\dots,n$.

  If we take $N_0$ to be at least the sum of the sizes of the nilpotent Jordan blocks $J_{n_0+1},\dots,J_n$, then for $N\geq N_0$ we have
  $$
  \phi^N =
  \begin{pmatrix}
    J_1^N & *** &\dots & \dots& \dots  &\dots & *** \\
    0 & J_2^N & \dots &\dots&\dots & \dots& *** \\
    \vdots &\vdots & \ddots &\vdots \\
    0 & 0 & \dots & J_{n_0}^N & *** & \dots & ***\\
    0 & 0 & \dots & 0 & 0 & \dots & 0\\
    \vdots &\vdots & \ddots &\vdots \\
    0 & 0 & \dots & 0 & 0 & \dots & 0
  \end{pmatrix},
  $$
  where each nonzero $J_i^N$ has the units $\lambda_i^N$ on the diagonal.
  ($\mu(M)$ is an upper bound on the sum of the sizes of the nilpotent Jordan blocks, so the above holds for $N\geq \mu(M)$.)

@@

  So, all that needs to be seen is that $\psi:=\psi^N$ is surjective.
  To see this, note that after replacing the original generators for $M$ by the generators that put $\psi$ into the form above,  a generating set for $\im(\phi^N)$ is given by $f_i = \psi(m_i) = u_i\cdot e_i + \sum_{j<i } a_{ij} e_j$ for $i = 1,\dots,n-N_0$, where $u_i$ is a unit (explicitly, $u_i$ is a power of some nonzero $\lambda_{i'}\in k$).
  That is, a generating set is given by
  $$u_1 e_1,\quad  u_2 e_2+a_{21}e_1,\quad \dots ,\quad u_{n-n_0}e_{n-N_0}+a_{n-N_0,n-1}e_{n-N_0-1}+\dots + a_{N-n_0,1}e_1.$$
  Since the $u_i$ are units, the $R$-span of these generators is clearly the same as the $R$-span of $e_1,\dots,e_{n-N_0}$, and thus after applying $\psi$ we get back the $R$-span of $f_1,\dots,f_{n-N_0} $, which is $\im \psi^N$. Thus, $\phi: \im \psi^N \to \im \psi^N$ is surjective, and the result follows.
\end{proof}

@@

The above can of course produce the trivial direct sum decomposition: if the eigenvalues of $\phi$ are all nonzero, the proof yields that $\im \psi = M$ and $\coker \psi = 0$. The following corollary of the proof of \cref{lem:graded_algorithm} ensures that we can always assume $\phi$ has zero as an eigenvalue:

\begin{lem}
  The eigenvalues of $\phi$ lie in $k$.
  In particular,
  for any  $\phi \in \End M$, there is $\lambda \in k$ such that $\phi - \lambda \id_M \in \End M$ has zero as an eigenvalue.
\end{lem}

@@

\begin{algm}\label{alg:graded}
  Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
  \begin{enumerate}
  \item\label{item:End0} Take a random element $\phi_0$ of $[\End_R(M)]_0$, the degree zero part of $\End_R(M)$.
  \item For an eigenvalue $\mu $ of $\phi_0$, let $\phi=\phi_0-\mu \id_M$.
  \item If $\phi$ has at least two eigenvalues, let $\psi=\phi^{\mu(M)}$. By \Cref{lem:graded_algorithm}, $M\cong \im \psi \oplus \coker \psi$.  \item Repeat steps (1)--(3) for both $\im \psi$ and $\coker \psi$.
  \end{enumerate}
\end{algm}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The local case}\label{sec:local-alg}

@@

\begin{lem}\label{lem:jordan}
  Let $k$ be a finite field of characteristic $p$, and let $A$ be an endomorphism of a $k$-vector space such that all eigenvalues of $A$ are contained in $k$. If $\lambda$ is an eigenvalue of $A$, then some power of $A-\lambda I$ is idempotent.
  Furthermore, if $\lambda$ is not the only eigenvalue of $A$, then a  nonzero power of $A-\lambda I$ is idempotent.
\end{lem}
\begin{proof}
  Since all eigenvalues of $A$ are contained in $k$, we can without loss of generality put $A$ in Jordan canonical form, with each Jordan block having the form
  \[ r_i\Biggl\{
  \underbrace{\begin{pmatrix}
      \lambda_i & 1 & 0 & \dots & 0 \\
      0 & \lambda_i & 1  & \dots & 0 \\
      \vdots & & & & \vdots \\
      0 & 0 & 0 & \dots & \lambda_i
  \end{pmatrix}}_{r_i} \]
  with each $\lambda_i$ an eigenvalue of $A$.
  In this basis, $A-\lambda I$ will be block-diagonal with blocks of form
  \[ \begin{pmatrix}
    \lambda_i-\lambda & 1 & 0 & \dots & 0 \\
    0 & \lambda_i-\lambda & 1 & \dots & 0 \\
    \vdots & & & & \vdots \\
    0 & 0 & 0 & \dots & \lambda_i-\lambda
  \end{pmatrix}. \]
  Set $\mu_i=\lambda_i-\lambda$. Then for any $l\geq 1$, the $l$-th power $(A-\lambda I)^l$ is block-diagonal with blocks of the form
  \[ \begin{pmatrix}
    \mu_i^l & \binom{l}1\mu_i^{l-1} & \binom{l}2\mu_i^{l-2} & \dots & \binom{l}{r_i}\mu_i^{l-r_i} \\
    0 & \mu_i^l & \binom{l}1\mu_i^{l-1} & \dots & \binom{l}{r_i-1}\mu_i^{l-r_i+1} \\
    \vdots & & & & \vdots \\
    0 & 0 & 0 & \dots & \mu_i^l
  \end{pmatrix}. \]
  If we choose $l > r_i$ and $l$ a power of $p$, say $p^{e_0}$, then all non-diagonal terms will vanish, so all blocks will have the form
  \[ \begin{pmatrix}
    \mu_i^l & 0 & 0 & \dots & 0 \\
    0 & \mu_i^l & 0 & \dots & 0 \\
    \vdots & & & & \vdots \\
    0 & 0 & 0 & \dots & \mu_i^l
  \end{pmatrix}. \]
  Finally, since $k\subset \FF_{p^e}$ for some $e$, if we choose $l$ to be divisible also by $p^e-1$, we have that $$\mu_i^{l-1} = (\mu_i^{p^{e}-1})^{l/(p^{e}-1)} =1.$$
  In particular, taking $l=p^{e_0}(p^e-1)$,
  $(A-\lambda I)^{p^{e_0}(p^e-1)}$ is a diagonal matrix with diagonal entries 1 or 0, hence idempotent. Note moreover that if some $\lambda_i\neq \lambda$, then $(A-\lambda I)^l$ is not the zero matrix.
\end{proof}

@@

\begin{algm}\label{alg:local}
  Given a finitely generated graded $R$-module $M$, the indecomposable summands of $M$ are computed as follows.
  \begin{enumerate}
    %In the graded case, take a general element $A_0$ of $[\End_R(M)]_0$, the degree zero part of $\End_R(M)$; in the local case,
  \item Take a general element $A_0$ of $\End_R(M) \setminus \m \End_R(M)$.
  \item Consider the resulting endomorphism $A$ of the $k$-vector space $M/\m M$ and find the eigenvalues of $A$.
  \item If $A$ has at least two eigenvalues, choose one eigenvalue $\lambda$, and compute a sufficiently high power $B=A^{p^{e_0}(p^{e}-1)}$ of $A$ (with the power explicitly as in the proof of \Cref{lem:jordan}). This power will be a nonzero idempotent, and thus produce a splitting of $M$ as $\im B \oplus \coker B$ by \Cref{lem:idemp}.
  \item Repeat steps (1)--(3) for both $\im B$ and $\coker B$.
  \end{enumerate}
\end{algm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Termination of the algorithms}

@@

\begin{lem}\label{lem:term}
  Assume that either $(R,\m)$ is a graded ring with $R_0=k$, or $(R,\m)$ is a local ring with $k=R/\m$ a field of positive characteristic, and that $M$ is a finitely generated $R$-module (homogeneous in the graded case).
  If $M$ has an endomorphism with at least two distinct eigenvalues modulo $\m$, then $M$ is not indecomposable.
\end{lem}

@@

\begin{lem}\label{lem:distinct}
  Retain the notation of \Cref{lem:term}.
  \begin{itemize}
  \item If $R$ is graded and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero has at least two distinct eigenvalues (necessarily in $k$)
  \item If $R$ is local and $M$ is not indecomposable, then a general endomorphism of $M$ of degree zero reduces to an endomorphism of $M/\m M$ with at least two distinct eigenvalues.
  \end{itemize}
\end{lem}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Decomposing coherent sheaves}\label{sec:coherent}

@@

\begin{dfn}
  For $f\in \End_{\OO_X}(E)$, taking top exterior powers yields the map
  $$ \bigwedge^{\rank E} f: \det E  \to \det E, $$
  where $\det E := \bigwedge^{\rank E} E$.
  Since $\End_{\OO_X}(\det E) = H^0(\OO_X)=k$, $\bigwedge^{\rank E} f$ is multiplication by $\lambda\in k$; we write $\det f$ for this scalar $\lambda$.
\end{dfn}

@@

\begin{dfn}
  $\lambda \in k$ is an \defn{eigenvalue} of $f \in \End_{\OO_X}(E)$ if $\det(f-\lambda \id_E)= 0$. In other words, the eigenvalues of $f$ are the zeroes of the univariate polynomial $\det(f-t\id _E)$ with coefficients in $k$.
\end{dfn}
